\begin{frontmatter}
% frontpage and declaration are automatically handled here by the class
\end{frontmatter}
\cleardoublepage

% ------------------------------------------
% Switch to Roman numbering
% ------------------------------------------
\pagenumbering{roman}

% ------------------------------------------
% Abstract
% ------------------------------------------
\cleardoublepage
\begin{abstract}
Deep generative models, such as Variational Autoencoders (VAEs), are widely used in autonomous driving for learning compact latent representations of visual data. However, like other deep neural networks, VAEs and associated classifiers are often treated as black-box models, making it difficult to understand the reasoning behind their predictions. This lack of transparency raises concerns about trust and reliability, particularly in safety-critical applications such as autonomous vehicle decision-making.

One promising approach to improving model interpretability is through counterfactual explanations (CEs), which identify minimal changes required to alter a model’s output. Counterfactuals provide human-interpretable insights by highlighting influential features, offering a compelling alternative to traditional attribution methods. In this thesis, I propose a novel counterfactual explanation framework that leverages a deep generative model (VAE) in combination with various feature masking strategies to generate CEs for black-box classifiers. By altering input images or their latent representations, the framework identifies minimal and semantically plausible perturbations that can change a classifier’s decision, revealing features with a causal influence on the model’s predictions.

To evaluate this approach, experiments are conducted using simulated driving data from the CARLA environment, covering both binary and multi-class decisions such as STOP, GO, LEFT, and RIGHT. The generated counterfactuals are assessed using quantitative metrics and visual inspection to evaluate plausibility, visual realism, and interpretability. Additionally, a human evaluation study was conducted through a web-based interface to measure user trust in the generated explanations. Results indicate that latent-space interventions particularly those guided by LIME with nearest-neighbor support as well as grid-based image masking, were consistently rated as the most trustworthy and informative by participants. This work contributes to the broader field of explainable AI by advancing interpretable model analysis for decision-making in autonomous systems.

The complete implementation of this framework is available at \href{https://github.com/darshandodamani/A-Counterfactual-Explanation-Approach-Using-Deep-Generative-Models} {github}.

\textbf{Keywords:}
Counterfactual Explanations, Explainable AI, Variational Autoencoders, Latent Space Manipulation, LIME, Human-Centered Evaluation, Masking, Trust in AI, Autonomous Driving, Black-box Classifier, Visual Interpretability.
\end{abstract}

\clearpage
\thispagestyle{empty}
\null
\clearpage

% ------------------------------------------
% Acknowledgements
% ------------------------------------------
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

I would like to express my deepest gratitude to my supervisor, Prof.\ Dr.\ Jan Oliver Ringert, for giving me the opportunity to carry out this master's thesis under his guidance. His invaluable mentorship, constructive feedback, and the freedom he gave me to explore independently were instrumental in shaping the direction and depth of this work. I would also like to thank Prof.\ Dr.\ Volker Rodehorst for kindly agreeing to serve as the second examiner and for his valuable feedback and support.

I am sincerely grateful to the Department of Software Engineering at the Bauhaus-Universität Weimar for providing the necessary computational resources, including high-performance GPUs and workstations. I would especially like to thank M.Sc. Benjamin Burse and M.Sc. Soaibuzzaman for their technical assistance in setting up the Linux system and configuring the GPU environment, which were essential for executing the experiments in this thesis.

Special thanks to my fellow master's students and colleagues who contributed to the user study conducted as part of this thesis. Your participation and insights made a meaningful impact on the human evaluation component of this work. I also thank my friends for their encouragement and support throughout this journey.

Above all, I extend my heartfelt appreciation to my family especially my mother, father, and brother for their unwavering belief in me. I am equally grateful to my girlfriend, Reeta, whose motivation, and constant support have been a pillar of strength during my academic journey.

Thank you all for being part of this milestone.



\clearpage
\thispagestyle{empty}
\null
\clearpage
% ------------------------------------------
% TOC, Figures, Tables
% ------------------------------------------
\tableofcontents
\cleardoublepage

\listoffigures
\cleardoublepage

\listoftables
\cleardoublepage

% ------------------------------------------
% Abbreviations
% ------------------------------------------
\cleardoublepage
\chapter*{Abbreviations}
\addcontentsline{toc}{chapter}{Abbreviations}
\input{text/Abbreviations.tex}
\cleardoublepage

% ------------------------------------------
% Nomenclature
% ------------------------------------------
\renewcommand{\nomname}{List of Symbols}
\addcontentsline{toc}{chapter}{List of Symbols} 
\input{text/Notations.tex}
\printnomenclature
\cleardoublepage

% ------------------------------------------
% Switch to Arabic numbering
% ------------------------------------------
\pagenumbering{arabic}

% Ensure main content starts on odd-numbered page
\cleardoublepage
\ifodd\value{page}\null\thispagestyle{empty}\newpage\fi

% ------------------------------------------
% Main Chapters
% ------------------------------------------
\chapter{Introduction} \label{Introduction}
\input{text/Introduction.tex}

\cleardoublepage
\ifodd\value{page}\null\thispagestyle{empty}\newpage\fi
\chapter{Background} \label{Background}
\input{text/Background.tex}

\cleardoublepage
\chapter{Methodology} \label{Methodology}
\input{text/Methodology.tex}

\cleardoublepage
\ifodd\value{page}\null\thispagestyle{empty}\newpage\fi
\chapter{Evaluation and Results} \label{Evaluation and Results}
\input{text/Evaluation and Results.tex}

\cleardoublepage
\ifodd\value{page}\null\thispagestyle{empty}\newpage\fi
\chapter{Related Work} \label{Related work}
\input{text/Related work.tex}

\cleardoublepage
\ifodd\value{page}\null\thispagestyle{empty}\newpage\fi
\chapter{Conclusion and Future Scope} \label{Conclusion and Future Scope}
\input{text/Conclusion and Future Scope.tex}

% ------------------------------------------
% Bibliography
% ------------------------------------------
\cleardoublepage
\printbibliography[heading=bibintoc]
\markboth{}{} % <-- clears "Bibliography" from headers

% ------------------------------------------
% Appendix
% ------------------------------------------
\cleardoublepage
\appendix

% Add this to fix appendix figure/table numbers (A.1, B.1, etc.)
\renewcommand{\thefigure}{\Alph{chapter}.\arabic{figure}}
\renewcommand{\thetable}{\Alph{chapter}.\arabic{table}}

% Appendix A
\refstepcounter{chapter}
\renewcommand{\thechapter}{A}
\chapter*{Appendix: Labeling Threshold Justification}
\markboth{Appendix: Labeling Threshold Justification}{}
\addcontentsline{toc}{chapter}{Appendix: Labeling Threshold Justification}
\label{appendix:labeling}
\input{text/Appendix_Dataset_Labeling}

% Appendix B
\refstepcounter{chapter}
\renewcommand{\thechapter}{B}
\chapter*{Appendix: Classifier Metrics}
\markboth{Appendix: Classifier Metrics}{}
\addcontentsline{toc}{chapter}{Appendix: Classifier Metrics}
\label{appendix:classifier}
\input{text/Appendix_Metrics.tex}

% Appendix C
\refstepcounter{chapter}
\renewcommand{\thechapter}{C}
\chapter*{Appendix: User Study Web Interface}
\markboth{Appendix: User Study Web Interface}{}
\addcontentsline{toc}{chapter}{Appendix: User Study Web Interface}
\label{appendix:webinterface}
\input{text/Appendix_WebInterface.tex}

% Appendix D
\refstepcounter{chapter}
\renewcommand{\thechapter}{D}
\chapter*{Appendix: Dataset Diversity Samples}
\markboth{Appendix: Dataset Diversity Samples}{}
\addcontentsline{toc}{chapter}{Appendix: Dataset Diversity Samples}
\label{appendix:diversity}
\input{text/Appendix_Diversity.tex}

% Appendix E
\refstepcounter{chapter}
\renewcommand{\thechapter}{E}
\chapter*{Appendix: Visual of Generated Counterfactuals}
\markboth{Appendix: Visual of Generated Counterfactuals}{}
\addcontentsline{toc}{chapter}{Appendix: Visual of Generated Counterfactuals}
\label{appendix:comparison}
\input{text/Appendix_Counterfactuals.tex}